---
title: "Getting Started with RSnowflake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RSnowflake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

RSnowflake is a DBI-compliant R interface to Snowflake that connects
directly to the Snowflake SQL API via HTTPS. It has no dependency on ODBC,
JDBC, or Python.

## Installation

```{r install}
# From GitHub (development version):
# install.packages("pak")
pak::pak("Snowflake-Labs/RSnowflake")
```

## Authentication

RSnowflake supports several authentication methods. The recommended
approach is to configure a profile in `~/.snowflake/connections.toml`,
the same file used by the Snowflake Python connector and Snowflake CLI.

### connections.toml (recommended)

```toml
[default]
account   = "myaccount"
user      = "myuser"
authenticator = "SNOWFLAKE_JWT"
private_key_path = "/path/to/rsa_key.p8"
database  = "MY_DB"
schema    = "PUBLIC"
warehouse = "MY_WH"
```

Then connect without any arguments:

```{r connect-toml}
library(DBI)
library(RSnowflake)

con <- dbConnect(Snowflake())
```

Or select a named profile:

```{r connect-named}
con <- dbConnect(Snowflake(), name = "my_profile")
```

### Programmatic Access Token (PAT)

Set the `SNOWFLAKE_PAT` environment variable, or pass the token directly:

```{r connect-pat}
con <- dbConnect(
  Snowflake(),
  account = "myaccount",
  token   = Sys.getenv("SNOWFLAKE_PAT")
)
```

### Snowflake Workspace Notebooks

Inside a Workspace Notebook, authentication is automatic -- the session
token is read from the environment:
```{r connect-workspace}
con <- dbConnect(Snowflake())
```

## Running Queries

```{r query}
# Simple query
df <- dbGetQuery(con, "SELECT * FROM my_table LIMIT 10")

# Parameterized query
df <- dbGetQuery(
  con,
  "SELECT * FROM users WHERE age > ?",
  params = list(21L)
)

# Execute DDL / DML (returns rows affected)
dbExecute(con, "CREATE TABLE test_tbl (id INT, name VARCHAR)")
dbExecute(con, "INSERT INTO test_tbl VALUES (1, 'Alice')")
```

## Writing and Reading Tables

```{r tables}
# Write a data.frame to Snowflake
dbWriteTable(con, "iris_copy", iris)

# Read it back
iris_back <- dbReadTable(con, "iris_copy")

# Append more rows
dbAppendTable(con, "iris_copy", iris[1:10, ])

# Clean up
dbRemoveTable(con, "iris_copy")
```

### Bulk Write Methods

RSnowflake automatically picks the fastest write path for your environment.
The `upload_method` option controls this:

| Method | How it works | Best for |
|---|---|---|
| `"auto"` (default) | Snowpark in Workspace, ADBC outside, literal for small data | Most users |
| `"snowpark"` | Python `write_pandas` via reticulate (internal SPCS path) | Workspace bulk writes |
| `"adbc"` | Arrow-native PUT + COPY INTO via Go driver | Local/external bulk writes |
| `"literal"` | SQL INSERT statements | Small data, maximum compatibility |

In **Workspace Notebooks**, `"auto"` routes bulk writes (above
`RSnowflake.bulk_write_threshold` cells) through the Python Snowpark session
via reticulate. This uses the internal SPCS network path and is ~10x faster
than ADBC for writes in Workspace (~2.5s vs ~21s for 50K rows).

**Outside Workspace**, `"auto"` routes to ADBC when available, which
outperforms literal INSERT for datasets above ~20K rows.

```{r upload-method}
# Force a specific method
options(RSnowflake.upload_method = "snowpark")
dbWriteTable(con, "big_table", large_df)

# Adjust the threshold (cells = rows * cols)
options(RSnowflake.bulk_write_threshold = 100000L)
```

## Identifier Case Handling

By default, RSnowflake uppercases table and column names in DDL/DML
operations (`dbWriteTable`, `dbCreateTable`, `dbReadTable`, etc.).
This matches the behavior of Snowflake's ODBC driver and of Snowflake
itself when identifiers are unquoted:

```{r case-upper}
dbWriteTable(con, "my_table", data.frame(id = 1, name = "Alice"))
dbListFields(con, "my_table")
#> [1] "ID" "NAME"
```

To preserve the original case of identifiers (useful for DBItest
compliance or when you need lowercase column names), set:

```{r case-preserve}
options(RSnowflake.identifier_case = "preserve")

dbWriteTable(con, "my_table", data.frame(id = 1, name = "Alice"), overwrite = TRUE)
dbListFields(con, "my_table")
#> [1] "id" "name"
```

**Note:** `dbQuoteIdentifier()` is not affected by this option -- it
always faithfully quotes the string you give it. The option only
controls the implicit casing applied in table-level operations.

## ADBC Acceleration (optional)

When the `adbcsnowflake` and `adbcdrivermanager` packages are installed,
RSnowflake can use the ADBC (Arrow Database Connectivity) backend for
Arrow-native reads and PUT + COPY INTO bulk writes. This is particularly
beneficial for large datasets outside of Workspace Notebooks.

```{r adbc}
# Check if ADBC is available
has_adbc <- requireNamespace("adbcsnowflake", quietly = TRUE)
cat("ADBC available:", has_adbc, "\n")

# ADBC is used automatically in 'auto' mode for large writes
# Force it for reads:
options(RSnowflake.backend = "adbc")
df <- dbGetQuery(con, "SELECT * FROM big_table")
options(RSnowflake.backend = "auto")
```

See `WORKSPACE_ADBC.md` in the package for Workspace-specific ADBC
constraints (EAI network rules, PAT auth, stage host whitelisting).

## Arrow Interface

RSnowflake supports the DBI Arrow methods for compatibility with packages
and workflows that expect them. This requires the `nanoarrow` package.

```{r arrow}
stream <- dbGetQueryArrow(con, "SELECT * FROM my_table")
df <- as.data.frame(stream)
```

**Note:** The Snowflake SQL API v2 only returns JSON, so these Arrow methods
fetch data through the normal JSON path and convert to nanoarrow on the
client side. This provides interface compatibility but no performance
advantage over `dbGetQuery()`. For large result sets, `dbGetQuery()` is
the most direct path. Native server-side Arrow transport is planned for
a future release.

## dbplyr Integration

RSnowflake integrates with dbplyr so you can use dplyr verbs that are
translated to Snowflake SQL and executed lazily:

```{r dbplyr}
library(dplyr)

tbl(con, "my_table") |>
  filter(score > 90) |>
  select(name, score) |>
  arrange(desc(score)) |>
  collect()
```

## Transactions

**Note:** The Snowflake SQL API v2 is stateless per-request and does not
currently support session-based transactions. Calling `dbBegin()` will
raise an informative error. Session-based transaction support is planned
for a future release that uses the internal Snowflake protocol.

```{r transactions, eval = FALSE}
# Not yet supported via SQL API v2 -- included for reference:
dbBegin(con)
dbExecute(con, "INSERT INTO accounts VALUES (1, 100.00)")
dbCommit(con)
```

## Disconnecting

```{r disconnect}
dbDisconnect(con)
```
